---
title: "EggNOG taxonomic classification analysis"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(stringr)
library(knitr)

ggplotRegression <- function(fit){
ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5))) +
  theme(panel.background = element_blank(),
        panel.grid.major.y = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())
}

```

```{r load_data}

dat <- read.table(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/Performance_analyses/TreeSAPP_EggNOG_raw_classifications.tsv",
                  header=TRUE,
                  sep="\t", quote="")

dat <- dat %>% 
  separate(TrueLineage, c("t_root", "t_kingdom", "t_phylum", "t_class", "t_order", "t_family", "t_genus", "t_species"),
           sep = "; ",
           fill = "right",
           extra = "drop") %>%
  separate(AssignedLineage, c("a_root", "a_kingdom", "a_phylum", "a_class", "a_order", "a_family", "a_genus", "a_species"),
           sep = "; ",
           fill = "right",
           remove = FALSE) %>% 
  mutate(True.Phylum = paste(t_root, t_kingdom, t_phylum, sep="; ")) %>% 
  mutate(Assigned.Phylum = paste(a_root, a_kingdom, a_phylum, sep="; "))
```

I used TreeSAPP to compare the taxonomic assignments in EggNOG (v4.5) to those from reference packages built using PFam-curated sequences. What taxa was TreeSAPP able to classify?

```{r}
high_tax_counts <- dat %>%
  filter(!is.na(t_kingdom)) %>% 
  mutate(True.Phylum = gsub("Root; ", '', x = True.Phylum, perl = TRUE)) %>%
  count(Classified, True.Phylum, t_kingdom)

overall_freq <- high_tax_counts %>%
  group_by(True.Phylum) %>% 
  mutate(total = sum(n)) %>%
  ungroup() %>%
  select(True.Phylum, total) %>%
  unique() %>%
  mutate(freq = total/sum(total))

high_tax_counts <- merge(high_tax_counts, overall_freq, by="True.Phylum") %>% 
  mutate(True.Phylum = reorder(True.Phylum, freq)) %>% 
  filter(freq >= 0.005)

ggplot(high_tax_counts, aes(x=True.Phylum, y=n)) +
  geom_bar(stat="identity", size=1,
           aes(colour=t_kingdom,
               fill=Classified)) +
  scale_fill_brewer(palette = "Greys") +
  scale_colour_brewer(palette = "Paired") +
  guides(colour=guide_legend(title="Kingdom"),
         size=FALSE) +
  ylab("Reference sequences") +
  xlab("Reference Phylum") +
  theme(panel.background = element_blank(),
        panel.grid.major.y = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/EggNOG_taxonomic_classifications.png",
       width = 8, height = 6)
```

What happened to the Eukaryotic sequences?

```{r}

euk_tax <- dat %>% 
  filter(t_kingdom == "Eukaryota") %>% 
  mutate(Assigned.Phylum = gsub("; NA.*", '', x = Assigned.Phylum, perl = TRUE)) %>% 
  mutate(Assigned.Phylum = gsub("Root; ", '', x = Assigned.Phylum, perl = TRUE))

euk_tax_counts <- euk_tax %>%
  select(RefPkg, Classified, TaxDist, Assigned.Phylum) %>%
  filter(Classified == "True") %>% 
  group_by(Assigned.Phylum) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(Assigned.Phylum = reorder(Assigned.Phylum, freq)) %>% 
  filter(freq >= 0.001)

ggplot(euk_tax_counts, aes(x=Assigned.Phylum, y=freq)) +
  geom_bar(stat="identity") +
  ggtitle("Taxonomic assignments of Eukaryotic sequences in EggNOG") +
  ylab("Proportion") +
  xlab("Assigned Phylum") +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/EggNOG_euks.png",
       width = 8, height = 6)
```

Alright, apprently 47% the Eukaryotic sequences are being assigned to the Proteobacteria and only 45% are assigned to either Root, Bacteria, or Archaea. How deep are these Proteobacteria assignments?

Note: the contributions to these classifications are varied across reference packages - there are not just one or two reference packages causing these shallow classifications. I made a really ugly plot to validate this!

```{r}
euk_proteos <- euk_tax %>% 
  filter(a_phylum == "Proteobacteria") %>%
  mutate(AssignedLineage = gsub("Root; Bacteria; ", '', x = AssignedLineage, perl = TRUE)) %>%
  group_by(AssignedLineage) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  ungroup() %>% 
  filter(freq >= 0.01) %>% 
  mutate(AssignedLineage = reorder(AssignedLineage, freq))

ggplot(euk_proteos, aes(x=AssignedLineage, y=freq)) +
  coord_flip() +
  geom_bar(stat="identity") +
  ylab("Proportion") +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/EggNOG_proteobacteria_euks.png",
       width = 8, height = 4)
```


What were sequences classified as?

```{r taxa_profile}
assigned <- filter(dat, Classified == "True") %>% 
  mutate(Assigned.Class = paste(a_root, a_kingdom, a_phylum, a_class, sep="; ")) %>%
  mutate(True.Class = paste(t_root, t_kingdom, t_phylum, t_class, sep="; ")) %>% 
  select(TaxDist, RefPkg, True.Phylum, True.Class, Assigned.Phylum, Assigned.Class,
         t_kingdom,
         a_root, a_kingdom, a_phylum, a_class, a_order, a_family, a_genus, a_species)
assigned$TaxDist <- as.factor(assigned$TaxDist)

n_species <- nrow(filter(assigned, !is.na(a_species)))
n_genus <- nrow(filter(assigned, !is.na(a_genus)))
n_family <- nrow(filter(assigned, !is.na(a_family)))
n_order <- nrow(filter(assigned, !is.na(a_order)))
n_class <- nrow(filter(assigned, !is.na(a_class)))
n_phylum <- nrow(filter(assigned, !is.na(a_phylum)))
n_kingdom <- nrow(filter(assigned, !is.na(a_kingdom)))
n_root <- nrow(filter(assigned, !is.na(a_root)))

rank_counts <- data.frame(Rank = c("Root", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
                          Count = c(n_root, n_kingdom, n_phylum, n_class, n_order, n_family, n_genus, n_species))

kable(rank_counts, format = "html")

assigned_census <- assigned %>% 
  count(Assigned.Phylum, TaxDist, sort = T, name = "Count") %>% 
  mutate(Assigned.Phylum = gsub("; NA", '', x = Assigned.Phylum, perl = TRUE)) %>%
  mutate(Assigned.Phylum = reorder(Assigned.Phylum, Count))

ggplot(filter(assigned_census, Count > 0.01*max(Count)), aes(x=Assigned.Phylum, y=Count, fill=TaxDist)) +
  geom_histogram(stat = "identity") +
  scale_fill_brewer(palette = "Reds") +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/TreeSAPP_EggNOG_assigned.png",
       width = 8, height = 4)
```

How do the abundances compare to each other?

```{r correlation}
assigned_counts <- assigned %>% 
  count(Assigned.Class, sort = T, name = "Count") %>% 
  mutate(log_n = log(Count))
true_counts <- assigned %>%
  count(True.Class, sort = T, name = "Count") %>% 
  mutate(log_n = log(Count))

count_comparison <- merge(true_counts, assigned_counts,
                          by.x = "True.Class", by.y = "Assigned.Class")

ggplotRegression(lm(Count.y ~ Count.x, data = count_comparison))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/TreeSAPP_EggNOG_abund_corr.png",
       width = 5, height = 5)
ggplotRegression(lm(log_n.y ~ log_n.x, data = count_comparison))

```

How does this compare to GraftM?

```{r}
graftm_dat <- read.table(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/Performance_analyses/GraftM_EggNOG_raw_classifications.tsv",
                  header=TRUE,
                  sep="\t", quote="")
assigned <- filter(graftm_dat, Classified == "True") %>%
  separate(TrueLineage, c("t_root", "t_kingdom", "t_phylum", "t_class", "t_order", "t_family", "t_genus", "t_species"),
           sep = "; ",
           fill = "right",
           extra = "drop") %>%
  separate(AssignedLineage, c("a_root", "a_kingdom", "a_phylum", "a_class", "a_order", "a_family", "a_genus", "a_species"),
           sep = "; ",
           fill = "right",
           remove = FALSE) %>% 
  mutate(True.Phylum = paste(t_root, t_kingdom, t_phylum, sep="; ")) %>% 
  mutate(Assigned.Phylum = paste(a_root, a_kingdom, a_phylum, sep="; ")) %>% 
  mutate(Assigned.Class = paste(a_root, a_kingdom, a_phylum, a_class, sep="; ")) %>%
  mutate(True.Class = paste(t_root, t_kingdom, t_phylum, t_class, sep="; ")) %>% 
  select(TaxDist, RefPkg, True.Phylum, True.Class, Assigned.Phylum, Assigned.Class,
         t_kingdom,
         a_root, a_kingdom, a_phylum, a_class, a_order, a_family, a_genus, a_species)
assigned$TaxDist <- as.factor(assigned$TaxDist)

assigned_census <- assigned %>% 
  count(Assigned.Phylum, TaxDist, sort = T, name = "Count") %>% 
  mutate(Assigned.Phylum = gsub("; NA", '', x = Assigned.Phylum, perl = TRUE)) %>%
  mutate(Assigned.Phylum = reorder(Assigned.Phylum, Count))

ggplot(filter(assigned_census, Count > 0.01*max(Count)), aes(x=Assigned.Phylum, y=Count, fill=TaxDist)) +
  geom_histogram(stat = "identity") +
  scale_fill_brewer(palette = "Reds") +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "#bdbdbd", linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/GraftM_EggNOG_assigned.png",
       width = 8, height = 4)

assigned_counts <- assigned %>% 
  count(Assigned.Class, sort = T, name = "Count") %>% 
  mutate(log_n = log(Count))
true_counts <- assigned %>%
  count(True.Class, sort = T, name = "Count") %>% 
  mutate(log_n = log(Count))

count_comparison <- merge(true_counts, assigned_counts,
                          by.x = "True.Class", by.y = "Assigned.Class")

ggplotRegression(lm(Count.y ~ Count.x, data = count_comparison))
ggsave(file="~/Bioinformatics/Hallam_projects/TreeSAPP_manuscript/manuscript/figures/GraftM_EggNOG_abund_corr.png",
       width = 5, height = 5)


```